---
title: "Report"
author: "Abrham Birru"
date: "12/24/2020"
output:
  pdf_document: default
  html_document: default
---
#################################################
# MovieLens Rating Project Code 
################################################

## Introduction
 This data analysis report is prepared as a part of HarvardX Data Science Capstone (HarvardX: PH125.9x) Project.The given dataset(movielens) has been analysed by using different tools and techniques. 

 In this project, I will combine several machine learning strategies to construct a movie recommendation system based on the "MovieLens" dataset.

## Goal
  The goal is to predict movie ratings, and evaluate the accuracy of the predicted model from the given code the dataset called  "edx" was split into the training and validation sets.

## Data Loading
 The code is provided in the edx capstone project module:
 The following lines of code will create training and validation sets

##########################################################
# Create edx set, validation set (final hold-out test set)
##########################################################

 Note: this process could take a couple of minutes

```{r}
if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")
if(!require(data.table)) install.packages("data.table", repos = "http://cran.us.r-project.org")

library(tidyverse)
library(caret)
library(data.table)

# MovieLens 10M dataset:
# https://grouplens.org/datasets/movielens/10m/
# http://files.grouplens.org/datasets/movielens/ml-10m.zip

dl <- tempfile()
download.file("http://files.grouplens.org/datasets/movielens/ml-10m.zip", dl)

ratings <- fread(text = gsub("::", "\t", readLines(unzip(dl, "ml-10M100K/ratings.dat"))),
                 col.names = c("userId", "movieId", "rating", "timestamp"))

movies <- str_split_fixed(readLines(unzip(dl, "ml-10M100K/movies.dat")), "\\::", 3)
colnames(movies) <- c("movieId", "title", "genres")

movies <- as.data.frame(movies) %>% mutate(movieId = as.numeric(levels(movieId))[movieId],
                                           title = as.character(title),
                                           genres = as.character(genres))


movielens <- left_join(ratings, movies, by = "movieId")

# Validation set will be 10% of MovieLens data
set.seed(1, sample.kind="Rounding")
test_index <- createDataPartition(y = movielens$rating, times = 1, p = 0.1, list = FALSE)
edx <- movielens[-test_index,]
temp <- movielens[test_index,]

# Make sure userId and movieId in validation set are also in edx set
validation <- temp %>% 
  semi_join(edx, by = "movieId") %>%
  semi_join(edx, by = "userId")

# Add rows removed from validation set back into edx set
removed <- anti_join(temp, validation)
edx <- rbind(edx, removed)
rm(dl, ratings, movies, test_index, temp, movielens, removed)

```

## Exploring The Data

```{r}
head(edx)
head(validation)
str(edx)
str(validation)
names(edx)
names(validation)
```
# summary of unique movies and users

```{r}
summary(edx)
```
# Number of unique movies and users in dataset 

```{r}
edx %>%
  summarize(number_of_users = n_distinct(userId), 
            number_of_movies = n_distinct(movieId))
```
# Plot Rating distribution

```{r}
edx %>%
  ggplot(aes(rating)) +
  geom_histogram(binwidth = 0.25, color = "blue") +
  scale_x_discrete(limits = c(seq(0.5,5,0.5))) +
  scale_y_continuous(breaks = c(seq(0, 3000000, 500000))) +
  ggtitle("Rating distribution")
```
# Make a plot of number of ratings per movie in edx dataset

```{r}
edx %>%
  count(movieId) %>%
  ggplot(aes(n)) +
  geom_histogram(bins = 30, color = "blue") +
  scale_x_log10() +
  xlab("Number of ratings") +
  ylab("Number of movies") +
  ggtitle("Number of ratings per movie")
```
# only once rated 20 movies table

```{r}
edx %>%
  group_by(movieId) %>%
  summarize(count = n()) %>%
  filter(count == 1) %>%
  left_join(edx, by = "movieId") %>%
  group_by(title) %>%
  summarize(rating = rating, n_rating = count) %>%
  slice(1:20) %>%
  knitr::kable()
```
# Make a plot of mean movie ratings given by users

```{r}
edx %>%
  group_by(userId) %>%
  filter(n() >= 100) %>%
  summarize(b_u = mean(rating)) %>%
  ggplot(aes(b_u)) +
  geom_histogram(bins = 30, color = "blue") +
  xlab("Mean rating") +
  ylab("Number of users") +
  ggtitle("Mean movie ratings given by users") +
  scale_x_discrete(limits = c(seq(0.5,5,0.5))) +
  theme_light()
```
# Make a plot of number of ratings given by users

```{r}
edx %>%
  count(userId) %>%
  ggplot(aes(n)) +
  geom_histogram(bins = 30, color = "blue") +
  scale_x_log10() +
  xlab("Number of ratings") + 
  ylab("Number of users") +
  ggtitle("Number of ratings given by users")
```
## Modelling

## Model-1

# Let's take the average(mean)
 This model is developed by using average rating (mean) to train the model and predict the movie rating in the validation model.

```{r}
average <- mean(edx$rating)
average
```
# Initiate RMSE and store based on simple prediction

```{r}
naive_rmse <- RMSE(validation$rating, average)
naive_rmse

rmse_data <- data_frame(method = "Average Movie Rating Model", RMSE = naive_rmse)
rmse_data %>% knitr::kable()
```
## Model-2

 In this model the Movies are used to examine the predictive efficiency
 Simple model taking into account the movie effect b_i
 Subtract the rating minus the mean for each rating the movie received
 Plot number of movies with the computed b_i

```{r}
movie_avgs <- edx %>%
  group_by(movieId) %>%
  summarize(b_i = mean(rating - average))
movie_avgs %>% qplot(b_i, geom ="histogram", bins = 10, data = ., color = I("blue"),
                     ylab = "Number of movies", main = "Number of movies with the computed b_i")

```
# Store rmse data results 

```{r}
predicted_ratings <- average +  validation %>%
  left_join(movie_avgs, by='movieId') %>%
  pull(b_i)
model_1_rmse <- RMSE(predicted_ratings, validation$rating)
rmse_data <- bind_rows(rmse_data,
                          data_frame(method="Movie Effect Model",  
                                     RMSE = model_1_rmse ))
```
# Finally Check rmse data results

```{r}
rmse_data %>% knitr::kable()
```

## Model-3
# Here we will model the Movie-User effects

```{r}
user_avgs<- edx %>% 
  left_join(movie_avgs, by='movieId') %>%
  group_by(userId) %>%
  filter(n() >= 100) %>%
  summarize(b_u = mean(rating - average - b_i))
user_avgs%>% qplot(b_u, geom ="histogram", bins = 30, data = ., color = I("blue"))
```

# User averages group by user id
```{r}
user_avgs <- edx %>%
  left_join(movie_avgs, by='movieId') %>%
  group_by(userId) %>%
  summarize(b_u = mean(rating - average - b_i))
```

# Test and save rmse data results 

```{r}
predicted_ratings <- validation%>%
  left_join(movie_avgs, by='movieId') %>%
  left_join(user_avgs, by='userId') %>%
  mutate(pred = average + b_i + b_u) %>%
  pull(pred)

model_2_rmse <- RMSE(predicted_ratings, validation$rating)
rmse_data <- bind_rows(rmse_data,
                          data_frame(method="Movie-User Effect Model",  
                                     RMSE = model_2_rmse))
```
# Check rmse data result
```{r}
rmse_data %>% knitr::kable()
```

## Model-4
# Here Cross validation is used to select optimum value of lambda.It is a regularized model.
```{r}
lambdas <- seq(0, 10, 0.25)
```

# For each lambda,find b_i & b_u, and followed by rating prediction and testing

```{r}
rmses <- sapply(lambdas, function(l){
  average <- mean(edx$rating)
  
  b_i <- edx %>% 
    group_by(movieId) %>%
    summarize(b_i = sum(rating - average)/(n()+l))
  
  b_u <- edx %>% 
    left_join(b_i, by="movieId") %>%
    group_by(userId) %>%
    summarize(b_u = sum(rating - b_i - average)/(n()+l))
  
  predicted_ratings <- 
    validation %>% 
    left_join(b_i, by = "movieId") %>%
    left_join(b_u, by = "userId") %>%
    mutate(pred = average + b_i + b_u) %>%
    pull(pred)
  
  return(RMSE(predicted_ratings, validation$rating))
})

```

# Make a Plot rmses vs lambdas to select the optimal lambda   

```{r}
qplot(lambdas, rmses)  
```

# The optimal lambda  

```{r}
lambda <- lambdas[which.min(rmses)]
lambda
```

# Test and save results  
```{r}
rmse_data <- bind_rows(rmse_data,
                          data_frame(method="Regularized Movie and User Effect Model",  
                                     RMSE = min(rmses)))
```
# Result    

```{r}
rmse_data %>% knitr::kable()
```

# Conclusion

We have built a machine learning algorithm to predict movie ratings with a given MovieLens dataset.The regularized model including the effect of user has lower RMSE value(0.8648170) which reduced the error further.We can further improve the RMSE by utilizing age,genre,year and so on.

## Appendix

```{r}
print("Operating System:")
version
```

